---
title: Aural landscapes of DuPage County
#output: rmarkdown::pdf_document
author: Andrew L. Hipp (ahipp@mortonarb.org)
date: 28 August 2021
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aural landscapes of DuPage County}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Running the code
First, let's grab my most recent observations:
```{r}
library(auralLandscapes)
require(openxlsx)
require(magrittr)
a <- read.xlsx('https://github.com/andrew-hipp/auralLandscapes/blob/main/data/auralObservations.xlsx?raw=true',
            detectDates=T)
## use this if you are using decimal time:
a$time <- a$time * 24
```

The data are organized by taxon, time, date, with additional optional fields that are currently ignored:

```{r}
names(a)
```

At this point, you only have to have the three columns `taxon`, `date`, and `time` in your spreadsheet. Any others are gravy. If you have your own data in an Excel spreadsheet in xlsx format, you can read it in using `read.xlsx` with `detectDates = T`, and then multiply the decimal times by 24 to get hours that make sense. This is a bit of a kludge, and other formats don't currently work so well. Note two bits of formatting fussiness:

 * dates have to be in YYYY-MM-DD format (e.g., 2021-08-20)
 * times are in HH:MM 24-hour format (e.g., 08:50 or 21:45)

At the outset, it might be useful to investigate how many taxa you have observed:

```{r}
a$taxon %>% unique %>% sort
a$taxon %>% unique %>% sort %>% length
```

... and probably double-check for orthographic errors (e.g., including both "cardinal" and "cardinals", or both "white-throated sparrow" and "white throated sparrow"). These can be corrected manually or on the fly; I would recommend cleaning up your data first b/f plotting rather than creating code to clean it on the fly.

Then, consider which observations you have the most of:

```{r}
a$taxon %>% table %>% sort(decreasing = TRUE) %>% head(20)
```

Once you've done some of this preliminary work, the question is what aspects of your aural landscape you want to explore.

## Looking at observations
First, the herps:
```{r, fig.width=8, fig.height = 4, out.width="100%"}
aur.herps <- plotAuralLandscape(a, taxa = 'american toad|peeper|chorus',
  dateRange=1:11, legPos = c(0.9, 0.82))
```

Then, robins and cardinals:
```{r, fig.width=8, fig.height = 4, out.width="100%"}
aur.robCard <- plotAuralLandscape(a, taxa = 'robin|cardinal', dateRange=c(1:11))
```

the leading edge of a few harbingers of spring:
```{r, fig.width=8, fig.height = 4, out.width="100%"}
aur.harbingers <-
  plotAuralLandscape(a,
    taxa = 'blackbird|peeper|chorus|song sparrow|black-throated|woodcock',
    dateRange = 2:5,
    legPos = NA, colorAlpha = 1.0)
```

How about the morning songs of very early spring:
```{r, fig.width=8, fig.height = 4, out.width="100%"}
aur.springMornings <- plotAuralLandscape(a,
  taxa = 'robin|dove|chorus|peeper|woodcock|cardinal|song sparrow',
  dateRange = 1:4, timeRange = c(4:6),
  legPos = NA, colorAlpha = 1.0)
```
